On the adhesivity properties of equivalence graphs\thanks{???????}

### 2025-01-31 icgt/main.tex

@@

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}
\def\E{\textbf {\textup{E}}}
\def\T{\textbf {\textup{1}}}
\def\A{\textbf {\textup{A}}}

%categorie varie
\newcommand{\catname}[1]{\textbf{\textup{#1}}}
\newcommand{\lab}{\catname{LHyp}}
\newcommand{\hyp}{\catname{Hyp}}
\newcommand{\hyps}{\catname{Hyp}_{\Sigma}}
\newcommand{\gr}{\textbf{\textup{Graph}}}
\newcommand{\dgr}{\catname{SGraph}}
\newcommand{\dg}{\catname{DAG}}
\newcommand{\rt}{\mathsf{dcl_s}}
\newcommand{\rta}{\mathsf{dcl}}
\newcommand{\rtd}{\mathsf{dcl_{d}}}
\newcommand{\slice}[2]{(\catname{#1}\downarrow{#2})}
\newcommand{\tg}[0]{\catname{TG}_{\Sigma}}
\newcommand{\teg}[0]{\catname{TeGr}_{\Sigma}}
\newcommand{\sv}[0]{\mathsf{Sieves}}
\newcommand{\mono}[1]{\mathsf{Mon}(\catname{#1})}
\newcommand{\mo}[1]{{#1}_\mathsf{Mon}}
\newcommand{\pro}{\mathsf{prod}}
\newcommand{\spro}{\mathsf{ps}}
\newcommand{\prol}{\mathsf{lprod}}
\newcommand{\pred}[1]{{\downarrow}#1}
\newcommand{\colim}[0]{\mathrm{colim}}
\newcommand{\cod}{\mathsf{cod}}
\renewcommand{\sp}{\mathsf{sp}}
\renewcommand{\sup}{\mathsf{sup}}
\newcommand{\cow}[1]{\mathsf{cwd}({#1})}
\renewcommand{\inf}{\mathsf{inf}}
\newcommand{\dom}{\mathsf{dom}}
\newcommand{\dwnarrow}{\downarrow \hspace{-2pt}}
\newcommand{\Dwnarrow}{\Downarrow \hspace{-2pt}}

@@

\todo{Va riscritta tutta con xy}

@@

\begin{theorem}\todo{controllare meglio}
	\label{thm:slice-functors}
	Let $\X$ be an $\mathcal{M}$-adhesive category. Given an object $X$
	the category $\X/X$ is $\mathcal{M}/X$-adhesive with
	$\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
	\mathcal{M}\}$. Similarly, $X/\X$ is $X/\mathcal{M}$-adhesive with
	$X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}$.

	Moreover for every small category $\Y$, the category $\X^\Y$ of
	functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
	$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
	\mathcal{M} \text{ for every } Y\in \Y\}$.
\end{theorem}

We can list various examples of $\mathcal{M}$-adhesive categories (see
\cite{castelnovo2023thesis,CastelnovoGM22,lack2006toposes}).

@@

\begin{proposition}\label{prop:kermono}
	An arrow $m\colon M\to X$ is mono if and only if $(M, id_M, id_M)$ is a kernel pair for it.
\end{proposition}

\begin{proof}
    To prove the ``if'' part of the statement, let $f, g: A \to M$ be such that $m\circ f = m\circ g$, and consider the following situation.
    \[
        \begin{tikzcd}[row sep=26, column sep = 26]
        A \arrow[drr, bend left=30, "f"] \arrow[ddr, bend right=30,"g" swap] \arrow[dr, dashed, "u"] & & \\
        & M  \arrow[d, "{id_M}"] \arrow[r, "{id_M}"] & M \arrow[d, "m"] \\
        & M  \arrow[r, "m" swap] & X
        \end{tikzcd}
    \]
    For the universal property of pullbacks, we have that $$f  =  id_M \circ u =  g$$
    Hence, $m$ is mono.

    Conversely, if $m$ is mono, then, we have that
    \begin{align*}
        m \circ f = m \circ g   &\Rightarrow    f = g \\
                                &\Rightarrow    f \circ id_M = g\circ id_M
    \end{align*}
    Hence, $f$ is the unique arrow that makes the commutative square illustrated above a pushout.
\end{proof}

@@

\begin{corollary}\label{cor:kermono}
	$(K_f, \pi_f^1, \pi_f^2)$ is a kernel pair for $f\colon X\to Y$ if and only if, for each mono $m\colon Y\to Z$, $(K_f, \pi_f^1, \pi_f^2)$ is a kernel pair also for $m\circ f$.
\end{corollary}
\begin{proof}
    It is enough to see that, by \Cref{lemma:pullback_lemma} and \Cref{prop:kermono} the outer boundary of the following square is a pullback.
        \[\begin{tikzcd}[row sep=13 pt, column sep=13 pt]
    	{K_f} && X && X \\
    	\\
    	X && Y && Y \\
    	\\
    	X && Y && Z
    	\arrow["{\pi_f^2}", from=1-1, to=1-3]
    	\arrow["{\pi_f^2}"', from=1-1, to=3-1]
    	\arrow["{id_X}", from=1-3, to=1-5]
    	\arrow["f", from=1-3, to=3-3]
    	\arrow["f", from=1-5, to=3-5]
    	\arrow["f", from=3-1, to=3-3]
    	\arrow["{id_X}"', from=3-1, to=5-1]
    	\arrow["{id_Y}", from=3-3, to=3-5]
    	\arrow["{id_Y}", from=3-3, to=5-3]
    	\arrow["m", from=3-5, to=5-5]
    	\arrow["f"', from=5-1, to=5-3]
    	\arrow["m"', from=5-3, to=5-5]
    \end{tikzcd}\]
    The leftward part of the statement follows by definition of monomorphism an \Cref{lemma:pullback_lemma}.
\end{proof}

@@

\subsection{Factorization systems}
\todo{Inserire qualche cosa su sist. di fattorizazzione e M-adesività - devo pensare a come mettere la questione senza prendere troppo spazio}
\todo{Di questo possiamo sbattercene perché tanto siamo su $\Set$}

@@

\appendix
\section{Some properties of comma categories}
In this section we will briefly recall the definition of the comma category associated to two functors and some of its properties.
\begin{definition}\index{category!comma -}
	Let $L\colon \A\to \X$ and  $R\colon \B\rightarrow \X$ be two functors with the same codomain, the \emph{comma category} $\comma{L}{R}$ is the category in which
	\begin{itemize}
		\item objects are triples $(A, B, f)$ with $A\in \A$, $B\in \B$, and $f\colon L(A)\rightarrow R(B)$; 
		\item a morphism $(A, B, f)\rightarrow (A', B', g)$ is a pair $(h, k)$ with $h\colon A\rightarrow A'$ in $\A$ and $k\colon B\rightarrow B'$ in $\B$ such that the following diagram commutes
		\[\xymatrix@C=25pt{L(A) \ar[r]^{L(h)} \ar[d]_{f} & L(A') \ar[d]^{g}\\ R(B) \ar[r]_{R(k)}& R(B')}\]
	\end{itemize}
\end{definition} 
We have two forgetful functors 	$U_L\colon \comma{L}{R}\to \A$ and $U_R\colon \comma{L}{R} \to \B$ given, respectively by
\[
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{A}{h}{A'}
\end{split}\quad 
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{B}{k}{B'}
\end{split}
\]

@@

We can notice another useful fact,  showing that in some cases we can guarantee  the existence of a left adjoint to $U_R$.

@@

A straightforward application of \cref{lim} now yields the following.
\begin{corollary}\label{cor:pbapp}
	If $\X$ has pullbacks, then for every object $X$, the slice $\X/X$ has pullbacks too.
\end{corollary}

@@

\begin{definition}
	Let $(\X, \otimes, I)$ be a monoidal category. An \emph{monoid} is a triple $(X, m, e)$ made by an object $X$ equipped with arrows $m\colon X\otimes X\to X$ and $e\colon I\to X$ making the following diagram commute.

	\[\]

	A \emph{morphism of monoids} between $(X, m, e)$ and $(X', m', e')$ is an arrow $f\colon X\to X'$ fitting in the diagram below.

	\[\]
Monoids with their morphisms form a category $\mono{X}$, such category has an obvious forgetful functor $\mo{U}\colon\mono{X}\to X$ sending $(X, m, e)$ to $X$ and a morphism to itself.
\end{definition}

We are interested in provide sufficient conditions to guarantee the existence of a left adjoint to $\mo{U}$. The idea is to mimic the construction of the \emph{kleene star} functor $\Set\to \Set$ \cite{bibid}, i.e.~ the functor sending $X$ to the set of finite words over it.

@@

\begin{theorem}
	Let $(\X, \otimes, I )$ be a countably distributive monoidal category. Then $\mo{U}$ has a left adjoint sending $\mo{F}$ to $\sum_{n\in \mathbb{N}} X^{\otimes n}$.
\end{theorem}

As a concluding step, we ask for a compatibility property between countable coproducts and pullbacks, to assure that $\mo{F}$ preserves pullbacks.

\begin{definition}
	Let $\X$ be a category and suppose that $(X, \{j_i\}_{i\in I})$ is a coproduct. We say that $(X, \{j_i\}_{i\in I})$ is \emph{stable under pullbacks} if, given an arrow $f\colon Y\to X$ such that for every $i\in I$ the pullback square below exists,  the cocone $(Y, \{k_i\}_{i\in I})$ is a coproduct.
	\[\xymatrix{Y_i \ar[r]^{k_i} \ar[d]_{p_i} & Y \ar[d]^{f}\\ X_i\ar[r]_{j_i} & X}\]

\end{definition}

@@

\begin{proposition}
	Let $(\X, \otimes, I)$ be a countably distributive monoidal category and suppose that countable  coproducts in $\X$ are stable under pullbacks. Then the functor $\mo{F}\colon \X\to \mono{X}$ preserves pullbacks.
\end{proposition}
\begin{proof}
	contenuto...
\end{proof}

>>>>

\newcommand{\lgh}{\mathsf{lg}}

@@

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}
\def\E{\textbf {\textup{E}}}
\def\T{\textbf {\textup{1}}}
\def\A{\textbf {\textup{A}}}
\def\M{\mathcal{M}}
%categorie varie
\newcommand{\catname}[1]{\textbf{\textup{#1}}}
\newcommand{\lab}{\catname{LHyp}}
\newcommand{\hyp}{\catname{Hyp}}
\newcommand{\hyps}{\catname{Hyp}_{\Sigma}}
\newcommand{\gr}{\textbf{\textup{Graph}}}
\newcommand{\dgr}{\catname{SGraph}}
\newcommand{\dg}{\catname{DAG}}
\newcommand{\rt}{\mathsf{dcl_s}}
\newcommand{\rta}{\mathsf{dcl}}
\newcommand{\rtd}{\mathsf{dcl_{d}}}
\newcommand{\slice}[2]{(\catname{#1}\downarrow{#2})}
\newcommand{\tg}[0]{\catname{TG}_{\Sigma}}
\newcommand{\teg}[0]{\catname{TeGr}_{\Sigma}}
\newcommand{\sv}[0]{\mathsf{Sieves}}
\newcommand{\mono}[1]{\mathsf{Mon}(\catname{#1})}
\newcommand{\mo}[1]{{#1}_\mathsf{Mon}}
\newcommand{\pro}{\mathsf{prod}}
\newcommand{\spro}{\mathsf{ps}}
\newcommand{\prol}{\mathsf{lprod}}
\newcommand{\pred}[1]{{\downarrow}#1}
\newcommand{\colim}[0]{\mathrm{colim}}
\newcommand{\cod}{\mathsf{cod}}
\renewcommand{\sp}{\mathsf{sp}}
\renewcommand{\sup}{\mathsf{sup}}
\newcommand{\cow}[1]{\mathsf{cwd}({#1})}
\renewcommand{\inf}{\mathsf{inf}}
\newcommand{\dom}{\mathsf{dom}}
\newcommand{\dwnarrow}{\downarrow \hspace{-2pt}}
\newcommand{\Dwnarrow}{\Downarrow \hspace{-2pt}}

@@

@@

\begin{theorem}
	\label{thm:slice-functors}
	Let $\X$ be an $\mathcal{M}$-adhesive category, then the following hold true:
	\begin{enumerate}
		\item if $L\colon \X\to $
		\item for every object $X$
		the categories $\X/X$  and $X/X$ are, respectively, $\mathcal{M}/X$-adhesive and $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \qquad X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item given a  full and faithful $I\colon \Y\to \X$ and a class $\mathcal{M}'$ of arrows in $\Y$ stable under pullbacks, containing all isomorphism
	\end{enumerate} 
\end{theorem}

We will briefly list som examples of $\mathcal{M}$-adhesive categories.

@@

\begin{proposition}\label{prop:kermono}
	An arrow $m\colon M\to X$ is mono if and only if $(M, MM, \id{M})$ is a kernel pair for it.
\end{proposition}

\begin{proof}
    To prove the ``if'' part of the statement, let $f, g: A \to M$ be such that $m\circ f = m\circ g$, and consider the following situation.
    \[
        \begin{tikzcd}[row sep=26, column sep = 26]
        A \arrow[drr, bend left=30, "f"] \arrow[ddr, bend right=30,"g" swap] \arrow[dr, dashed, "u"] & & \\
        & M  \arrow[d, "{\id{M}}"] \arrow[r, "{\id{M}}"] & M \arrow[d, "m"] \\
        & M  \arrow[r, "m" swap] & X
        \end{tikzcd}
    \]
    For the universal property of pullbacks, we have that $$f  =  \id{M} \circ u =  g$$
    Hence, $m$ is mono.

    Conversely, if $m$ is mono, then, we have that
    \begin{align*}
        m \circ f = m \circ g   &\Rightarrow    f = g \\
                                &\Rightarrow    f \circ \id{M} = g\circ \id{M}
    \end{align*}
    Hence, $f$ is the unique arrow that makes the commutative square illustrated above a pushout.
\end{proof}

@@

\begin{corollary}\label{cor:kermono}
	$(K_f, \pi_f^1, \pi_f^2)$ is a kernel pair for $f\colon X\to Y$ if and only if, for each mono $m\colon Y\to Z$, $(K_f, \pi_f^1, \pi_f^2)$ is a kernel pair also for $m\circ f$.
\end{corollary}
\begin{proof}
    It is enough to see that, by \Cref{lemma:pullback_lemma} and \Cref{prop:kermono} the outer boundary of the following square is a pullback.
        \[\begin{tikzcd}[row sep=13 pt, column sep=13 pt]
    	{K_f} && X && X \\
    	\\
    	X && Y && Y \\
    	\\
    	X && Y && Z
    	\arrow["{\pi_f^2}", from=1-1, to=1-3]
    	\arrow["{\pi_f^2}"', from=1-1, to=3-1]
    	\arrow["{id{X}}", from=1-3, to=1-5]
    	\arrow["f", from=1-3, to=3-3]
    	\arrow["f", from=1-5, to=3-5]
    	\arrow["f", from=3-1, to=3-3]
    	\arrow["{id_X}"', from=3-1, to=5-1]
    	\arrow["{id_Y}", from=3-3, to=3-5]
    	\arrow["{id_Y}", from=3-3, to=5-3]
    	\arrow["m", from=3-5, to=5-5]
    	\arrow["f"', from=5-1, to=5-3]
    	\arrow["m"', from=5-3, to=5-5]
    \end{tikzcd}\]
    The leftward part of the statement follows by definition of monomorphism an \Cref{lemma:pullback_lemma}.
\end{proof}

@@

\input{./fact_sys}

@@

\appendix
\section{Some properties of comma categories}
In this section we will briefly recall the definition of the comma category \cite{mac2013categories} associated to two functors and some of its properties.
\begin{definition}\index{category!comma -}
	Let $L\colon \A\to \X$ and  $R\colon \B\rightarrow \X$ be two functors with the same codomain, the \emph{comma category} $\comma{L}{R}$ is the category in which
	\begin{itemize}
		\item objects are triples $(A, B, f)$ with $A\in \A$, $B\in \B$, and $f\colon L(A)\rightarrow R(B)$; 
		\item a morphism $(A, B, f)\rightarrow (A', B', g)$ is a pair $(h, k)$ with $h\colon A\rightarrow A'$ in $\A$ and $k\colon B\rightarrow B'$ in $\B$ such that the following diagram commutes
		\[\xymatrix@C=25pt{L(A) \ar[r]^{L(h)} \ar[d]_{f} & L(A') \ar[d]^{g}\\ R(B) \ar[r]_{R(k)}& R(B')}\]
	\end{itemize}
\end{definition} 
We have two forgetful functors 	$U_L\colon \comma{L}{R}\to \A$ and $U_R\colon \comma{L}{R} \to \B$ given, respectively by
\[
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{A}{h}{A'}
\end{split}\quad 
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{B}{k}{B'}
\end{split}
\]

@@

\begin{lemma}\label{colim}
	Let $L\colon \A\to \X$ and $R\colon \B\to \X$ be functors and $F\colon \D\to\comma{L}{R}$ be a diagram such that $L$ preserves colimits along $U_L\circ F$. Then the family $\{U_L, U_R\}$ jointly creates colimits of $F$ (see \cite{castelnovo2023thesis,castelnovo2022newcriterionmathcalmmathcalnadhesivity}).
\end{lemma}
\begin{proof}
	Suppose that $U_L\circ F$ and $U_R\circ F$ have colimiting cocones  $\left(A, \{a_D\}_{D\in \D}\right)$ and $\left(B, \{b_D\}_{D\in \D}\right)$ respectively. By hypothesis $\left(L(A), \{L\left(a_D\right)\}_{D\in \D}\right)$ is colimiting for $L\circ U_L\circ F$. Now, if we define
	\[F(D):=(A_D, B_D, f_D)\]
	then we have arrows $R(a_i)\circ f_D\colon L(A_D)\to R(B)$ that forms a cocone on $L\circ U_L\circ F$: if $d\colon D\to D'$ is an arrow in $\D$ then $F(d)$ is an arrow in $\comma{L}{R}$ and so
	\begin{align*}
		R\left(b_{D'}\right)\circ f_{D'}\circ L(U_L(F(d)))&=R\left(b_{D'}\right)\circ R\left(U_R\left(F(d)\right)\right)\circ f_D\\&=R\left(b_{D'}\circ U_R\left(F(d)\right)\right)\circ f_D\\&=R\left(b_D\right)\circ f_D
	\end{align*}
	Thus there exists $f\colon L(A)\rightarrow R(B)$ such that
	\[\xymatrix@C=35pt{ L(A_D) \ar[r]^{L(a_D)} \ar[d]_{f_D}& L(A) \ar[d]^{f}\\ R(B_D) \ar[r]_{R(b_D)} &R(B)}\]
	Notice that $f$ is the unique arrow in $\X$ wich makes $\left(a_D, b_D\right)$ an arrow $\left(A_D, B_D, f_D\right)\to \left(A, B, f\right)$ of $\comma{L}{R}$. If we show that $\left((A, B, f), \left\{(a_D, b_D)\right\}_{D\in \D}\right)$ is colimiting for $F$ we are done.

	First of all, let us show that it is a cocone. Given $d\colon D\to D'$ in $\D$ we have:
	\begin{align*}
		\left(a_{D'}, b_{D'}\right)\circ F(d)&=	\left(a_{D'}, b_{D'}\right)\circ \left(U_L(F(d)), U_R(F(d))\right)\\&=\left( a_{D'}\circ U_L(F(d)),  b_{D'}\circ U_R(F(d)) \right)\\&=\left(a_D, b_D\right)
	\end{align*}
	For the colimiting property, let $\left((X, Y, g), \left\{\left(x_D, y_D\right)\right\}_{D\in \D}\right)$ be another cocone on $F$. In particular $\left(X, \left\{x_D\right\}_{D\in \D}\right)$ and $\left(Y, \left\{y_D\right\}_{D\in \D}\right)$ are cocones on $U_L\circ F$ and $U_R\circ F$ respectively, so we have uniquely determined arrows $x\colon A\rightarrow X$ and $y\colon B\rightarrow Y$ such that 
	\[x\circ a_D= x_D \qquad y\circ b_D= y_D\]
	Let us show that $(x,y)$ is an arrow of $\comma{L}{R}$. Given $D\in \D$ we have
	\begin{align*}
		R(y)\circ f\circ L(a_D)&=R(y)\circ R\left(b_D\right)\circ f_D\\&=R\left(y\circ b_D\right)\circ f_D\\&=R\left(y_D\right)\circ f_D\\&= g\circ L\left(x_D\right)\\&=g\circ L\left(x\circ a_D\right)\\&=g\circ L(x)\circ L\left(a_D\right)
	\end{align*}
	from which it follows  that the following diagram commutes.
	\[\xymatrix@C=35pt{L(A)\ar[d]_{f} \ar[r]^{L(x)} & X \ar[d]^{g}\\ R(B) \ar[r]_{R(y)}& Y}\]
	This shows that $\left((A, B, f), \left\{(a_D, b_D)\right\}_{D\in \D}\right)$ is colimiting for $F$ and the thesis follows.
\end{proof}

\Cref{prop:dual} and \Cref{colim} now yields the following.
\begin{corollary}\label{lim} The family $\{U_L, U_R\}$ jointly creates limits along every diagram $F\colon \D\to \comma{L}{R}$ such that $R$ preserves the limit of $U_R\circ I$.
\end{corollary}

We can use \cref{lim} to characterize monos in comma categories. 
\begin{corollary}\label{cor:mono}
	If $R$ preserves pullbacks then an arrow $(h,k)$ in $\comma{L}{R}$ is mono if and only if both $h$ and $k$ are monomorphisms.
\end{corollary}
\begin{proof}
	$(\Rightarrow)$  If $(h,k)\colon (A,B,f)\to (A', B', g)$ is a mono then the following square is a pullback in $\comma{L}{R}$
	\[\xymatrix@C=50pt{(A, B, f)  \ar[r]^{\id{(A,B,f)}} \ar[d]_{\id{(A,B,f)}}& (A, B, f)  \ar[d]^{(h,k)} \\ (A, B, f)  \ar[r]_{(h,k)}& (A', B', g) }\]
	Using \cref{lim} we deduce that the following two squares are pullbacks in $\A$ and $\B$.
	\[\xymatrix{A \ar[r]^{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]^{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ A \ar[r]_{h} & A' & B \ar[r]_{k} & B'}\]
	From which it follows that $h$ and $k$ are monos.

	\smallskip\noindent 
	$(\Leftarrow)$ Since $h$ and $k$ are monos then we have two pullback squares 	
	\[\xymatrix{A \ar[r]^{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]^{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ A \ar[r]_{h} & A' & B \ar[r]_{k} & B'}\]
	By \cref{lim} this implies that 
	\[\xymatrix@C=50pt{(A, B, f)  \ar[r]^{\id{(A,B,f)}} \ar[d]_{\id{(A,B,f)}}& (A, B, f)  \ar[d]^{(h,k)} \\ (A, B, f)  \ar[r]_{(h,k)}& (A', B', g) }\]
	is a pullback in $\comma{L}{R}$ and we are done.
\end{proof}

We end this section pointing out another useful fact,  showing that in some cases we can guarantee  the existence of a left adjoint to $U_R$.

@@

A straightforward application of \Cref{lim} now yields the following.
\begin{corollary}\label{cor:pbapp}
	If $\X$ has pullbacks, then for every object $X$, the slice $\X/X$ has pullbacks too.
\end{corollary}

@@

\begin{definition}
	Let $(\X, \otimes, I)$ be a monoidal category. An \emph{monoid} is a triple $(X, m, e)$ made by an object $X$ equipped with arrows $m\colon X\otimes X\to X$ and $e\colon I\to X$ making the following diagrams commute, where $\alpha$ is the associator and $\lambda$ and $\rho$ the two unitors.

	\[\xymatrix@C=35pt{(X\otimes X) \ar[r]^{\alpha_{X, X, X}} \ar[dd]_{m\otimes \id{X}}\otimes X & X\otimes (X\otimes X) \ar[d]^{\id{X}\otimes m} & X \ar[r]^{\rho^{-1}_X}  \ar[d]_{\lambda^{-1}_X} \ar[ddr]_{\id{X}}& X\otimes I \ar[d]^{\id{X}\otimes e}\\& X\otimes X \ar[d]^{m}  & I\otimes X \ar[d]_{e\otimes \id{X}} & X\otimes X \ar[d]^{m} \\X\otimes X  \ar[r]_{m}& X & X\otimes X \ar[r]_{m} & X}\]

	A \emph{morphism of monoids} between $(X, m, e)$ and $(X', m', e')$ is an arrow $f\colon X\to X'$ fitting in the diagrams below.

	\[\xymatrix{X\otimes X \ar[r]^{f\otimes f} \ar[d]_{m}& Y\otimes Y \ar[d]^{m'} && I\ar[dr]^{e'} \ar[dl]_{e}\\ X\ar[r]_{f} & Y & X \ar[rr]_{f} && Y }\]
Monoids with their morphisms form a category $\mono{X}$, such category has an obvious forgetful functor $\mo{U}\colon\mono{X}\to X$ sending $(X, m, e)$ to $X$ and a morphism to itself.

If the functor $\mo{U}$ has a left adjoint $\mo{F}\colon \X\to \mono{\X}$, we will define the \emph{Kleene star} functor $(-)^\star\colon \X \to \X$ as $\mo{U}\circ \mo{F}$.
\end{definition}

We are interested in provide sufficient conditions to guarantee the existence of a left adjoint to $\mo{U}$. The idea is to mimic the construction of the usual functor $\Set\to \Set$ \cite{sakarovitch2009elements} sending $X$ to the set of finite words over it.

@@

\begin{theorem}
	Let $(\X, \otimes, I )$ be a countably distributive monoidal category. Then $\mo{U}\colon \mono{X}\to \X$ has a left adjoint  $\mo{F}\colon \X\to \mono{X}$ defined as
\[	\functor[l]{X}{f}{Y}
\functormapsto
\rfunctor{\sum_{i\in \mathbb{N}} X^{\otimes n}}{\sum_{i\in \mathbb{N}} f^{\otimes n}}{\sum_{i\in \mathbb{N}} Y^{\otimes n}}\]
\end{theorem}

\begin{remark}\label{rem:lg} Let $\X$ be a category with finite limits and suppose that $(\X, \times, 1)$ is countably distributive.  Then for every object $X$ in $\X$ we have a \emph{length homomorphism} $\lgh_X\colon \mo{F}(X)\to \mo{F}(1)$. The underlying morphism $X^\star \to 1^\star$ in $\X$ is simply $\sum_{n\in \mathbb{N}} (!_X)^{\otimes n}$.  Notice, moreover, that for every $f\colon X\to Y$ we have a commutative triangle as the one below.
	\[\xymatrix{X^\star \ar[rr]^{f^\star} \ar[dr]_{\lgh_X}&& Y^\star  \ar[dl]^{\lgh_Y}\\ & 1^\star }\]
\end{remark}

As a concluding step, we ask for a compatibility property between countable coproducts and pullbacks, to assure that $\mo{F}$ preserves pullbacks \cite{bibid}.

\begin{definition}
	Let $\X$ be a category, we say that $\X$ is \emph{countably extensive} if:
	\begin{enumerate}
		\item  for every coprojections $j_i \colon X_i\to \sum_{i\in I}^{X_i} $ in a countaple coproduct and arrow $f\colon Y\to \sum_{i\in I} X_i$, a pullback of $j_i$ along $f$ exists; 
		\item given a countable set $I$, a coproduct $(X, \{j_i\}_{i\in I})$, an arrow $f\colon Z\to X$ and a family of squares
		\[\xymatrix{Z_i \ar[r]^{k_i} \ar[d]_{f_i} & Z\ar[d]^{f} \\ X_i \ar[r]_{j_i} & X}\]
		then the above squares are all pullbacks if and only if $(Z, \{k_i\}_{i\in I})$ is a coproduct.
	\end{enumerate} 
\end{definition}

\begin{remark}\label{rem:pb}
Suppose that  $(\X, \times, 1)$ is a countably distributive monoidal category, with $\X$ countably extensive. Then the following squares are pullbacks for every $n\in \mathbb{N}$ and object $X$.
\[\xymatrix{X^{n} \ar[r]^-{j_n} \ar[d]_{!_{X^n}}& X^\star \ar[d]^{\lgh_X} & Y^{n} \ar[r]^-{k_n} \ar[d]_{!_{Y^n}}& Y^\star \ar[d]^{\lgh_Y}\\ 1 \ar[r]_{a_n}&  1^\star & 1 \ar[r]_{a_n}&  1^\star}\]
\end{remark}

An elementary, but quite important property of countably extensive category is the following.

\begin{proposition}\label{prop:inj}
Let $\X$ be a countably extensive category, then countable coproducts are disjoints and have monos as coprojections. 
\end{proposition}
\begin{proof}
Let $(X, \{j_i\}_{i\in I})$ be a coproduct for some countable set $I$. For every $i\in I$, consider the family ${k_{i, h}}_{h\in I}$ defined as
\[k_{i, h}:=\begin{cases}
	?_{X_{i}} & h\neq i\\
	\id{X_i} &h=i
\end{cases}\]
It is immediate to notice that $(X_{i}, \{k_{i,h}\}_{h\in I})$ is a countable coproduct. Thus the squares below, where $h\neq i$, are all pullbacks.
	\[\xymatrix{0 \ar[r]^{?_{X_h}} \ar[d]_{?_{X_h}}& X_h \ar[d]^{j_h} & X_i \ar[r]^{\id{X_i}} \ar[d]_{\id{X_i}}& X_i \ar[d]^{j_i}\\ X_i \ar[r]_{j_i} & X & X_i \ar[r]_{j_i} & X}\]
	The thesis now follows at once.
\end{proof}

@@

\begin{proposition}
	Let $\X$ be a category with finite limits such that the cartesian structure $(\X, \times, 1)$ is countably distributive. Suppose, moreover, that countable  coproducts in $\X$ are stable under pullbacks. Then the Kleene star $(-)^\star \colon \X\to \X$ preserves pullbacks.
\end{proposition}
\begin{proof}
 Suppose that the solid part of the diagram below is given, where  the inner square is the image of a pullback through $(-)^\star$.

 \[\xymatrix{T \ar@/^.4cm/[drr]^{t_1} \ar@/_.4cm/[ddr]_{t_2} \ar@{.>}[dr]^t\\ & P^\star  \ar[d]_{q^\star}\ar[r]^{p^\star}& X^\star \ar[d]^{f^\star} \\ & Y^\star\ar[r]_{g^\star} & Z^\star}\]

	For every $n\in \mathbb{N}$, consider the coprojection $k_n\circ Z^n\to Z^\star$ and let $T_n$ be the pullback below.
	\[\xymatrix{T_n \ar[r]^{p_n} \ar[d]_{q_n} & T  \ar[d]^{f^\star \circ t_1}\\
	Z^n \ar[r]_{k_n}& Z^\star}\]

	Notice that, by \Cref{rem:lg} and since $f^\star \circ t_1$ is equal to $g^\star \circ t_2$, we have
	\[
	\begin{split}
		\lgh_X\circ t_1 \circ p_n&=\lgh_Z\circ f^\star \circ t_1\circ p_n\\&=\lgh_Z\circ k_n\circ q_n\\&=a_n\circ !_{Z^n}\circ q_n\\&=a_n\circ !_{T_n}
	\end{split}\qquad \begin{split}
		\lgh_Y\circ t_2 \circ p_n&=\lgh_Z\circ g^\star \circ t_2\circ p_n\\&=\lgh_Z\circ k_n\circ q_n\\&=a_n\circ !_{Z^n}\circ q_n\\&=a_n\circ !_{T_n}
	\end{split}\]

	Thus in the diagram below the solid part commutes, yielding the dotted $x_n\colon T_n\to X^n$ and $y_n\colon T_n\to Y^n$.

	\[\xymatrix{T_n  \ar@/_.3cm/[ddr]_{!_{T_n}}\ar@{.>}[dr]^{x_n} \ar[r]^{p_n} & T \ar[dr]^{t_1}&& T_n \ar@/_.3cm/[ddr]_{!_{T_n}} \ar[r]^{p_n} \ar@{.>}[dr]^{y_n}& T \ar[dr]^{t_2}\\ &X^n   \ar[d]_{!_{X^n}}\ar[r]^{j_n}& X^\star \ar[d]^{\lgh_X}& & Y^n \ar[d]_{!_{Y^n}}  \ar[r]^{h_n}& Y^\star \ar[d]^{\lgh_Y}\\ & 1 \ar[r]_{a_n}& 1^\star && 1  \ar[r]_{a_n}& 1^\star}\]

	Now, limits commutes with limits, and, moreover, we have
	\begin{align*}	
	k_n\circ f^n\circ x_n & =f^\star \circ j_n\circ x^n\\&=f^\star \circ t_1\circ p_n\\&=g^\star \circ t_2\circ p_n\\&=g^\star \circ h_n\circ y_n\\&=k_n \circ g^n\circ y_n 
	\end{align*}
	so, by \Cref{prop:inj}, the solid part of the diagram below commutes an the inner square is a pullback, therefore we obtain the dotted arrow $b_n\colon T_n\to P^n $.
	\[\xymatrix{T_n \ar@{.>}[dr]^{b_n}  \ar@/^.4cm/[drr]^{x_n} \ar@/_.4cm/[ddr]_{y_n} \\ & P^n \ar[d]_{q^n} \ar[r]^{p^n}& X^n \ar[d]^{f^n}\\ &  Y^n  \ar[r]_{g^n}& Z^n}\]

	Since $\X$ is countably extensive, then $(T, \{p_n\}_{n\in \mathbb{N}})$ is a coproduct for the family $\{T_n\}_{n\in \mathbb{N}}$ and so we can define the arrow $t\colon T\to P^\star$ as $\sum_{n\in \mathbb{N}} b_n$
